---
title: "PRIDEC"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{PRIDEC}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup}
library(PRIDEC)

library(dplyr)
library(ggplot2)
```

The PRIDE-C package serves to facilitate the forecasting of infectious diseases using data from [DHIS2](https://dhis2.org/). It follows a workflow of 
1. Data processing
2. Model training
3. Forecasting

This demo will illustrate the process of forecasting a disease across multiple organisation units (e.g. administrative zones) three months into the future based on climatic variables.

# The demo data

The demo data is simulated malaria case data from Ifanadiana District, Madagascar collected at the level of the health center. It includes information data on monthly case counts, rainfall, temperature, socio-economic levels, elevation, mosquito net use, and geographic isolation. 

Generally, the value we want to predict is referred to as the `y_var` in the API calls, while the predictor variables are referred to as `pred_vars`.

It includes an associated dataset of spatial polygons representing each health facility catchment.

Below is a plot of the time series for four example health centers.

```{r}
data("demo_malaria")
data("demo_polygon")

demo_malaria |>
  filter(orgUnit %in% sample(unique(orgUnit), 4)) |>
  mutate(date = as.Date(paste0(period,"01"), format = "%Y%m%d")) |>
  ggplot(aes(x = date, y = n_case)) +
  geom_line() +
  facet_wrap(~orgUnit, scales = "free") +
  ylab("Number of cases")
```

# Data Processing

## `prep_data()`

The first step is to process and clean the data. The PRIDE-C forecasting workflow has four steps to prepare the data for the model:

1. **lag predictors**: in order to predict into the future, we must use lagged values of the predictor variables, so that the number of cases is predicted as a function of the variable at time `t-lag`
2. **normalize predictors**: all of the predictor variables are centered and scaled to help with model convergence
3. **Impute missing predictors**: missing values in the predictor variables are imputed via a seasonal imputation
4. **create spatial structure**: one of the model fitting approaches (INLA) uses the spatial structure of the data to help predict. This structure is created via a matrix that identifies the neighbors of each polygon.

The data preparation step is done via the `prep_data()` function. We will choose to lag all of the dynamic predictor variables and scale all of the predictor variables.

```{r}
prep_output <- prep_data(raw_data = demo_malaria,
                      y_var = "n_case",
                      lagged_vars = c("pev", "rain_mm", "temp_c"),
                      scaled_vars = c("pev", "rain_mm", "temp_c",
                                      "wealth_index", "elevation",
                                      "time_to_district"),
                      lag_n = 3,
                      graph_poly = demo_polygon)
```

`prep_data()` returns three elements in a list:

- `W_graph`: object containing the spatial structure of data required by INLA
- `data_prep`: data.frame of the cleaned and processed data
- `scale_factors`: data.frame of center and scaling factors used to normalize the predictor variables. This will be used when explorign variables later.

## `split_cv_rolling()`

Once the data is prepared, it needs to be split into multiple cross-validation sets (`cv_set`). Each `cv_set` contains an `analysis` and `assessment` dataset that will be used in model training. These `cv_set`s are created via rolling origin splits ([Tashman 2000](https://www.sciencedirect.com/science/article/abs/pii/S0169207000000650?via%3Dihub).) 

Each `cv_set` can be thought of as an individual test of the workflow's ability to forecast, where a model is fit to the `analysis` data following the specified `pred_vars` and model parameters and performance is assessed on the `assessment` data.Generally, the `analysis` data corresponds to a longer period of historical data used for training, while the `assessment` data corresponds to the actual forecast horizon relevant to the user.

We will use 5 years of `analysis` data to train the model, and 3 months of `assessment` data to test it. The function `split_cv_rolling` will return a list, where each element corresponds to one `cv_set`.

```{r}
splits_out <- split_cv_rolling(prep_output$data_prep,
                               month_analysis = 60,
                               month_assess = 3)
```

We can inspect a `cv_set` to better understand the structure of the data:

```{r}
lapply(splits_out[[24]], head)
```

# Model Training

## Modeling approaches

There are five modeling approaches built into the PRIDE-C forecasting workflow:

### naive

The naive model can be thought of as the null model with which to compare other forecasting approaches to. It creates forecasts equal to historical averages based on grouping variables, by default the month of the year and `orgUnit`. It does not use any predictor variables in predictions.

### arimax

The ARIMAX model is an ARIMA model fit with exogeneous variables fit individually to each `orgUnit`. The exogeneous variables must be dynamic and at the same temporal resolution at the case data. Each ARIMAX model is fit using the `forecast::auto.arima()` function. Initial models are fit using the full list of provided `pred_vars`, which are iteratively removed (beginning at the end of the list) until a model successfully converges. Data are log-transformed within the function to help with fitting, but forecasts are provided on the response-scale.
 
### glm_nb

A generalized linear model with a negative binomial distribution can be fit to the full dataset. It includes a variable to control for the `orgUnit` and month of the year, in addition to the specified predictor variables. It can be thought of as the naive model with exogenous predictors. Predictor variables can be either dynamic or constant within an `orgUnit`, allowing for the inclusion of socio-demographic or topographical variables. The model is fit via the `MASS::glm.nb()` function.
 
### ranger

A random forest quantile regression model is implemented via the [`ranger` package](https://cran.r-project.org/web/packages/ranger/index.html). This package was chosen for its ability to apply quantile regression, its speed, and high-performance with default hyper-parameters. Three hyper-parameters (`m.try`, `min.node.size`, and `num.trees`) can be tuned with the `tune_ranger()` function. Predictor variables can be either dynamic or constant within an `orgUnit`, allowing for the inclusion of socio-demographic or topographical variables. It includes a variable to control for the `orgUnit` and month of the year, in addition to the specified predictor variables. 

### inla

An approximate Bayesian model is fit via Integrated Nested Laplace Approximation via the [`inla` package](https://www.r-inla.org/). Note that this package is not available on CRAN and should be installed following these [installation instructions](https://www.r-inla.org/download-install). 

This is the most computationally intense model, as it includes both temporal and spatial random effect structures. The temporal random effects are included via cyclical random walk with an annual seasonality, nested within `orgUnit`. The spatial structure is defined by a Besag-York-Mollié (BYM) model. This is an autoregressive spatial model based on the neighborhood spatial structure of the `orgUnits`. Priors can be provided for both random effects via the `hyper_priors` argument to `fit_inla`. In the future, functionality will be added to tune these hyper-parameters.

# Training Workflow

Each modeling approach follows the same training workflow:

1. `tune_*`: tuning of hyperparameters (`ranger` only)
2. `fit_*`: fitting of models to multiple `cv_set`s
3. `eval_performance`: evaluation of models on `analysis` and `assessment` datasets
4. `inv_variables_*`: investigation of variable importance scores and counterfactual plots to visualize the association with predictor variables

This workflow can be launched for multiple approaches at once and returns a `quarto` report of model performance and variable investigation via the `train_models` function. 

