---
title: "Template for model validation for PRIDE-C workflow"
lang: en
format: 
  html:
    toc: true
    toc-location: right
    embed-resources: true
    page-layout: full
editor: source
execute:
  echo: false
  warning: false
  output: false
execute-dir: project
---

```{r}
#| eval: false

#to render via a test
# quarto render validation_report_template.qmd -M title:"PRIDE-C Forecast Report" --output forecast_report.html  

#NOTE: this always needs to be rendered from the output_dir used by `run_pridec_forecast`

```

```{r packages}
#| include: false

library(jsonlite)
library(skimr)
library(sf)
# library(PRIDEC)

library(ggplot2); theme_set(theme_bw())
library(tidyr)
library(paletteer)

library(dplyr)

```

```{r load data}
#| include: false
config <- jsonlite::fromJSON("config.json")
forecast_df <- jsonlite::fromJSON(txt = "forecast.json")$dataValues
input_set <- readRDS("input_data.RData")
analysis_data <- input_set$analysis
assess_data <- input_set$assessment
poly <- sf::st_read("polygon.geojson", quiet = TRUE)
```

# Model Configurations and Info


**Date Run:** `r Sys.Date()`

::: {.callout-note collapse=true appearance="simple"}

## Model Configurations

```{r}
#| output: true
config$model_configs$inla$W_orgUnit <- NULL
print(config)
```

:::

# Input Data

To investigate errors in input data, run the following in `R`:

```{r}
#| eval: false
#| echo: true
readRDS("output/input_data.RData")
```

## Disease Variable

The model is trained on the `r config$disease_dataElement` data source from the DHIS2 instance.

:::: {.panel-tabset}

### Completeness

```{r}
#| output: true
#| 
y_var <- analysis_data[[config$disease_dataElement]]
prop_missing <- mean(is.na(y_var))
prop_outliers <- mean(y_var>(mean(y_var, na.rm = TRUE) + 3*sqrt(var(y_var, na.rm = TRUE))), na.rm = TRUE)

print(data.frame("Proportion Missing" = round(prop_missing,4),
                 "Proportion 3 sd outlier" = round(prop_outliers,4)))

#add missing by month and orgUnit
```

### Data Trends

Each panel represents a randomly chosen `orgUnit`. The historical case counts are plotted for each `orgUnit` for the full analysis period.

```{r}
#| output: true
#| 
analysis_data |>
  filter(orgUnit %in% sample(unique(analysis_data$orgUnit), 9)) |>
  rename(yvar := config$disease_dataElement) |>
  ggplot(aes(x = date, y = yvar)) +
  geom_line() +
  ylab(config$disease_dataElement) +
  xlab("Date") +
  facet_wrap(~orgUnit)
```

::::

## Predictor Variables

The model includes the following predictor variables: 


```{r}
pred_vars <- config$pred_vars
pred_var_df <- select(assess_data, orgUnit, period, date, all_of(pred_vars))
```

```{r}
#| output: true
 print(pred_vars)
```


:::: {.panel-tabset}

### Completeness

```{r}
#| output: true

skimr::skim(select(pred_var_df, -date))
```

```{r}
#| output: true
missing_cols <- colSums(is.na(pred_var_df))
missing_cols <- missing_cols[missing_cols>0]

for(this_col in names(missing_cols)){
  this_subset_na <- select(pred_var_df, orgUnit, date, values = this_col) |>
    dplyr::summarise(prop_NA = mean(is.na(values)),
                  .by = "date") |>
    filter(prop_NA > 0)
  print(paste("Missing in", this_col, ":"))
  print(this_subset_na)
}
```

To investigate missing climate variables, look directly at the climate data imported from DHIS2:

```{r}
#| eval: false
#| echo: true
climate_data <- jsonlite::fromJSON(txt = "input/climate_data.json")$dataValues
```


### Histograms

```{r}
#| output: true

# TODO: add check to only plot numeric/continuous
pred_var_df |>
  dplyr::mutate(Comparison = ifelse(date >= (max(date) - lubridate::years(1)), 
                                    "Prior Year", "Prior 3 Years")) |>
  tidyr::pivot_longer(cols = all_of(pred_vars), names_to = "variable", values_to = "value") |>
  ggplot(aes(x = value)) +
  geom_density(aes(fill = Comparison), alpha = 0.5, position = "identity") +
  facet_wrap(~variable, scales = "free")
```


### Timeseries

Timeseries of each predictor variable. A random selection of 8 orgUnits are shown, with each color line corresponding to one orgUnit.

```{r}
#| output: true

pred_var_df |>
  filter(orgUnit %in% sample(unique(pred_var_df$orgUnit), 8)) |>
  tidyr::pivot_longer(cols = all_of(pred_vars), names_to = "variable", values_to = "value") |>
  ggplot(aes(x = date, y = value)) +
  geom_line(aes(color = orgUnit)) +
  paletteer::scale_color_paletteer_d("khroma::bright") +
  facet_wrap(~variable, scales = "free_y") +
  theme(legend.position = 'none')
  
```

:::: 

# Forecasts

Created forecast for:

- `r length(unique(forecast_df$orgUnit))` orgUnits
- time period spanning `r min(forecast_df$period)` thru `r max(forecast_df$period)`

## Completion

List of orgUnits without full forecast period (`r config$month_assess` months beginning `r config$forecast_start`):

```{r}
#| output: true

forecast_df |>
  filter(period>=config$forecast_start, endsWith(dataElement, "Avg")) |>
  summarise(num_forecast = sum(!is.na(value)),
            .by = "orgUnit") |>
  filter(num_forecast<config$month_assess)
```

There should be no orgUnits listed in this list. If there are, this may mean that some input data is missing.

## Value Validity

Forecasts should be:

- integers
- non-negative
- real numbers

```{r}
#| output: true
PRIDEC::validate_dhis_format(forecast_df)
```

All should be zero.

## Time Series

```{r}
#| output: true

forecast_df |>
  filter(orgUnit %in% sample(unique(orgUnit), 9)) |>
  mutate(forecast_period = ifelse(period>=config$forecast_start, "Forecast", "Historic")) |>
  mutate(date = as.Date(paste0(period, "01"), format = "%Y%m%d")) |>
  mutate(conf_ints = case_when(
    endsWith(dataElement, "LowCI") ~ "lowCI",
    endsWith(dataElement, "UppCI") ~ "uppCI",
    endsWith(dataElement, "Avg") ~ "avg"
  )) |>
  select(-dataElement) |>
  pivot_wider(names_from = conf_ints, values_from = value) |>
  ggplot(aes(x = date, color = forecast_period, fill = forecast_period)) +
  geom_ribbon(aes(ymin = lowCI, ymax = uppCI), alpha = 0.3, color = NA) +
  geom_line(aes(y = avg)) +
  facet_wrap(~orgUnit, scales = "free_y") +
  scale_color_manual(values = c("navyblue", "gray20")) +
  scale_fill_manual(values = c("navyblue", "gray20")) +
  theme(legend.position = "bottom")
```

## Historic Performance

```{r}
#| output: true

forecast_df |>
  filter(orgUnit %in% sample(unique(orgUnit), 9)) |>
  filter(period<config$forecast_start) |>
  mutate(conf_ints = case_when(
    endsWith(dataElement, "LowCI") ~ "lowCI",
    endsWith(dataElement, "UppCI") ~ "uppCI",
    endsWith(dataElement, "Avg") ~ "avg"
  )) |>
    select(-dataElement) |>
  pivot_wider(names_from = conf_ints, values_from = value) |>
  left_join(select(analysis_data, orgUnit, period, 
                   starts_with("pridec_historic")), by = c("orgUnit", "period")) |>
  rename(observed = starts_with("pridec_historic")) |>
    mutate(date = as.Date(paste0(period, "01"), format = "%Y%m%d")) |>
  ggplot(aes(x = date)) +
  geom_ribbon(aes(ymin = lowCI, ymax = uppCI), fill = "gray20", alpha = 0.3) +
  geom_point(aes(y =observed)) +
  geom_line(aes(y = avg), color = "gray20") +
  ylab("Cases") +
  xlab("Date") +
  facet_wrap(~orgUnit, scales = "free_y")
```


## Map

```{r}
map_preds <- forecast_df |>
  filter(period>=config$forecast_start, endsWith(dataElement, "Avg")) |>
  summarise(avg_forecast = round(mean(value)),
            .by = "orgUnit") |>
  right_join(poly, by = "orgUnit") |>
  sf::st_as_sf()
```

```{r}
#| output: true

ggplot() +
  geom_sf(data = map_preds, aes(fill = avg_forecast)) +
  scale_fill_viridis_c()
```

::: {.callout-note collapse=true appearance="simple"}

# R Session Info

```{r}
#| output: true
sessionInfo()
```

:::
